Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes

### 2020-05-07 Ben_Manuscripts/hdphmm/Draft.tex
Filling in missing parts of draft

\documentclass{article}

\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{subcaption}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath} % or simply amstext
\usepackage{amssymb}
\usepackage{siunitx}
\usepackage{booktabs}
\usepackage[export]{adjustbox}
\newcommand{\angstrom}{\textup{\AA}}
\newcommand{\colormap}{jet}  % colorbar to use
\usepackage{cleveref}
\usepackage{booktabs}
\usepackage{gensymb}
\usepackage{float}
\usepackage{xr}
\usepackage{setspace}
\doublespacing
%\DeclareUnicodeCharacter{03C9}{$\omega$}
\externaldocument[S-]{Supporting_Information}

\title{Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series 
       of Solute Trajectories in Nanostructured Membranes.}

\author{Benjamin J. Coscia, Christopher P. Calderon, Michael R. Shirts}

@@

@@

%  There is a need for highly selective membranes in order to perform efficient 
%  separations of components of complex aqueous streams.
%  %MRS2: add some more examples (When you get around to it).
%  \begin{itemize}
%    \item Organic micropollutants
%    \item Desalination and boric acid removal from seawater.
%    \item While many researchers focus on membrane permeability, we may be 
%    able to reduce costs of commercial nanofiltration and reverse osmosis with
%    higher selectivity.~\cite{werber_materials_2016}
%  \end{itemize}
%
%%  \noindent Amphiphilic molecules are capable of self-assembling into ordered nanostructures.
%%  \begin{itemize}
%%    \item 
%%  \end{itemize}
%
%  Lyotropic liquid crystals (LLC) are a class of amphiphilic molecules whose ordered phases
%  can be cross-linked into mechanically strong membranes capable of highly selective
%  separations.
%  \begin{itemize}
%    \item The shape of the LLC monomers and water content dictates the ordered phase 
%    that they form. There are two phases of particular interest for membrane applications.
%  	\item H\textsubscript{II} phase lyotropic liquid crystals are characterized by 
%  	hexagonally packed, straight, pores while the Q\textsubscript{I} phase consists of
%  	a tortuous network of 3D interconnected pores. 
%  	\item In both cases the pores are uniform in size with radii on the order of 1 nm
%    %MRS2: is it really a molecular weight cutoff?
%    %BJC2: is molecular size better? Obviously there are other factors like shape, but I don't want to get into that
%  	giving them a very strict molecular size cut-off.
%	\item Additionally, they have the potential to disrupt conventional membrane separation
%	techniques by being selective based not only on size and charge, but on chemical
%	functionality as well.
%	\item Their pores are lined with LLC monomer functional groups which can potentially
%	be designed to interact with solutes in a chemically-specific manner
%  \end{itemize}
%
%  There are limits to what we can learn from experiment about LLC membrane design.
%  \begin{itemize}
%    \item Experimental observables like permeability and selectivity allow us to speculate
%    about the molecular origins of separation processes.
%    %MRS2: not quite sure what you mean below?
%    \item This drives an empirical design approach which can potentially neglect key
%    interactions which influence selectivity.
%    \item LLC membranes have been shown to exhibit selectivities which cannot be fully 
%    explained by relatively simple macroscopic models. 
%  \end{itemize}
%  
%  Molecular Dynamics (MD) simulations can give us mechanistic insights with atomistic 
%  resolution so that we can intelligently design new membranes for solute-specific 
%  separations.
%  \begin{itemize}
%    \item In our previous work, we built a detailed atomistic model which we used
%    to understand the nanoscopic structure of an LLC Membrane.~\cite{coscia_understanding_2019}
%    \item We also used the model in order to gain a qualitative understanding of 
%    trapping mechanisms which lead to subdiffusive transport behavior.~\cite{coscia_chemically_2019}
%  \end{itemize}
%
%  Unfortunately, the timescales that we can simulate with MD are insufficient to be
%  able to make well-converged predictions of macroscopic transport properties 
%  traditionally used to characterize membranes in the lab.
%  \begin{itemize}
%    %MRS2: would be good to get this next point into the first line of a paragraph, since it's quite important. 
%    \item However, if we use descriptive stochastic models that can capture solute
%    dynamics, then we could project long timescale behavior in addition to gaining
%    a deeper understanding of solute behavior on short timescales.
%  \end{itemize}

  %%%%%%%%%%%%%%%%%%%%%%%% Thesis chapter starts here %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

  In our previous study, we designed two different approaches which used
  solute time series in order to parameterize stochastic models that could be used
  to project transport on much longer timescales. In our first approach we modeled 
  solute trajectories as subordinated fractional Brownian and L\'evy motion, called
  the anomalous diffusion (AD) model. We generated solute trajectories by generating
  a series of anti-correlated hops separated by random periods of entrapment drawn 
  from a power law distribution. Our second approach treated solute motion as a Markov
  state model with state-dependent dynamics, called the Markov state-dependent dynamical
  model (MSDDM). We parameterized the state transition probabilities between
  each of eight discrete states as well as the solute dynamics within each of these
  states. We generated stochastic trajectory realizations by drawing a state
  sequence based on the transition probability matrix and incorporating the state dynamics
  while solutes were trapped in each state.

  Although both models had reasonable success at predicting solute mean squared 
  displacements (MSDs) on MD simulation timescales, they had shortcomings.
  The MSDDM failed to reproduce the hopping and trapping behavior that
  characterizes solute center-of-mass trajectories in our MD simulations.
  The AD model did not suffer this qualitative shortcoming, but the persistent 
  curvature of the predicted MSD curves suggested that the model might 
  underestimate MSDs on long timescales. The formulation of both models required
  careful examination and characterization of the interactions and dynamics exhibit
  by MD trajectories which required considerable human effort.

  In this work we attempt to minimize human effort by applying the infinite hidden 
  Markov Model (IHMM)~\cite{beal_infinite_2002,calderon_inferring_2015}, a nonparameteric 
  Bayesian classification technique, in order to automatically detect and infer the 
  parameters of an unknown number of latent autoregressive (AR) modes present in solute
  center-of-mass time series trajectories.
  We aim to address two primary scientific questions with this work. First, we want
  to know if the IHMM algorithm can help researchers efficiently uncover underlying
  transport mechanisms which give rise to different solute dynamical behavior.
  Second, we want to evaluate the ability of our model to reproduce solute trajectories
  in a stochastic manner so that we can use it to predict macroscopic transport 
  properties which would otherwise require much longer MD simulations.

@@

  Because independent solute trajectories generated by MD simulations share 
  similar state dynamics, we can use the IHMM to generate stochastic trajectory 
  realizations which sample from all behavior exhibited by MD. We verify that our
  stochastic realizations reproduce the expected hopping and trapping behavior 
  exhibited by MD. We quantitatively validate these trajectories by comparing 
  their mean squared displacement (MSD) to those measured by MD.

@@

  %Based on the algorithms developed by Fox et al.~\cite{fox_sticky_2007}, 
  Using the IHMM framework, we estimated the most likely number and sequence of hidden
  states in each solute center-of-mass trajectory while simultaneously inferring VAR(1)
  parameters for each state and the overall state transition probability matrix, $T$.
  We created a python implementation of this process which we heavily adapted from
  the MATLAB code of Fox et al.~\cite{fox_sticky_2007} Parameter estimation is iterative. 
  Therefore, we looked for convergence of each entry in the $\mathbf{c}$ vector as well
  as the $A$ and $\Sigma$ matrices of Equation~\ref{eqn:var}. We detected equilibration 
  of the parameters using the module \texttt{pymbar.timeseries.detect\_equilibration}~\cite{chodera_simple_2016} on
  the time series of parameter estimates. Plots illustrating parameter convergence are 
  available in Section~\ref{S-section:convergence} of the Supporting Information. We refer the interested reader to 
  much more extensive descriptions of the inference and sampling procedures used to 
  estimate the VAR(1) parameters and the state sequence. ~\cite{beal_infinite_2002,teh_hierarchical_2006,van_gael_beam_2008,fox_nonparametric_2009,fox_bayesian_2010}

  We carefully applied the IHMM algorithm in a way which takes advantage of the
  system's cylindrical symmetry. There are a number of viable ways in which one 
  could choose to analyze these time series. We could use all 3 dimensions straight
  from the output trajectory. We could use just the $z$ dimensional coordinate 
  since we are primarily interested in through-plane membrane transport. Ultimately, 
  we chose to use a multi-step procedure that we believe adequately distinguishes 
  every type of distinct dynamical behavior exhibited by the solutes. The final
  parameterization of the model is in cylindrical coordinates since each pore is a
  cylinder where we expect solutes to exhibit radially symmetric dynamics. We 
  present a step-by-step demonstration of our procedure in Section~\ref{S-section:ihmm_procedure}
  of the Supporting Information.

@@

  We applied the IHMM to each of the 24 solute trajectories independently.
  Although the IHMM is capable of identifying an infinite number of states, 
  a Dirichlet process tends to exhibit a ``rich get richer" effect, favoring
  a fewer number of states.~\cite{dreyer_discovering_2011} By applying the algorithm to each trajectory 
  independently, we reduce the possibility of lumping together multiple 
  similar states which we would prefer to stay separated before clustering.
  Note that the states identified by the IHMM are heavily influenced by 
  the Gaussian prior placed on $\mathbf{c}$ in Equation~\ref{eqn:var}. 
  In Section~\ref{S-section:prior_guesses} of the Supporting Information we outline our method for 
  choosing reliable prior parameters. The entries of $A$ and $\Sigma$ do 
  not vary over a wide range, so the final parameters were relatively insensitive
  to their priors.

  We ran 2000 iterations of the IHMM procedure in order to arrive at converged 
  state sequences. We evaluated the convergence of the state sequence based on
  the convergence of the VAR parameters. The parameters of each states
  generally converge within 500--1000 iterations. However, the boundaries of 
  each state segment fluctuate which can lead to high variance in the $A$
  and $\Sigma$ parameters in segments with a relatively low number of emissions.
  In these cases, the mean in each dimension tends to be a more reliable indicator of
  convergence (see Figure~\ref{S-fig:convergence3d} of the Supporting Information).

  We defined the finalized parameters of each state as the average of the 
  parameters from each iteration recorded after the equilibration time point.
  We found the equilibration time point of each component of the $\mathbf{c}$ 
  vector as well as $A$ and $\Sigma$ matrices, and then used the longest 
  equilibration time of all dimensions as the equilibration time point.
%  Further details are available in Section~\ref{S-section:convergence} of
%  the Supporting Information.

@@

@@

  We reduced the parameter space used agglomerative clustering, a hierarchical
  clustering approach which uses a linkage criteria in order to successively merge
  similar clusters until a desired intracluster distance threshold or number of
  clusters is reached.~\cite{pedregosa_scikit-learn_2011} We used the Ward linkage criteria, which works to minimize 
  the sum of the squared differences within all clusters.~\cite{ward_hierarchical_1963}
    %MRS6: explain why Ward (make the cluster as minimally disperse?)
   	%BJC6: I'll add some plots to the SI and then write a better justification for
	% the paper. This is lower priority in mind for the thesis.
  We elected to choose the number of clusters rather than the distance threshold.
  For our data, non-parametric methods such as Bayesian Gaussian mixture models~\cite{pedregosa_scikit-learn_2011,gelman_bayesian_2013} tend
  to delocalize the clusters in parameter space (see Section~\ref{S-section:agglomerative}
  of the Supporting Information).

  We clustered based on the two diagonal entries of $A$ and $\Sigma$ as well as the
  radial means, $\mu_r$, for a total of 5 variables. One can choose alternative clustering
  features, such as the eigenvalues of the $A$ and $\Sigma$ matrices, but we chose their 
  diagonals because they can be easily translated to solute behavior. The diagonals of $A$
  describe the correlation between sequential hops while the diagonals of $\Sigma$ describe
  the size of the hops in each dimension. We assume that cross correlations in the parameters
  are negligible for clustering.

  We also chose to cluster the diagonals of each matrix and the radial means
  independently before combining the clusters to assign final state labels. If we divide
  the total states into $m$ clusters based on the diagonal entries of $A$, $n$ more clusters 
  based on the diagonal entries of $\Sigma$, and $k$ clusters based on $\mu_r$, then there
  are a total of $mnk$ possible cluster combinations. This allows slightly more flexibility
  in the total number of states compared to clustering on all five features at once, since 
  there can be up to, but not necessarily exactly, $mnk$ states.

  Choosing the number of clusters can be somewhat subjective so
  we attempted to add some structure to the selection process by following
  a set of qualitative and quantitative guidelines. We used the silhouette 
  test in order to score the quality of clustering as a function of the number 
  of clusters chosen.~\cite{kaufman_finding_2009} For our data, the silhouette test generally favors the 
  lowest number of clusters possible (see Section~\ref{S-section:nclusters} of
  the Supporting Information). However, choosing too few clusters tends to 
  not distinguish between visually obvious differences in 
  dynamic behavior. This results in finalized parameter sets that are averages of
  distinct behavior which presents further problems with the predictive modeling
  that we discuss later on. We aimed to maintain the highest silhouette score,
  and thus lowest number of clusters, possible while verifying that visually 
  distinct states stayed separated. Based on these guidelines, we decided to 
  group the $A$ and $\Sigma$ parameters into five clusters each and the radial
  means into three clusters. Of the 75 distinct states allowed by this formulation,
  the solutes in this study showed behavior from 32--36 distinct clusters. While
  this may seem like a high number of states, many are sampled infrequently.

  We remapped the state sequence based on the cluster assignments and generated a
  state transition probability matrix, $T$. The IHMM algorithm also produces an estimate
  of $T$, but since we fixed the state sequence, we decided to explicitly calculate $T$ 
  by counting the number of transitions between states compiled across all trajectories.

  We used the parameter inference component of the IHMM algorithm in order to 
  infer $A$ and $\Sigma$ of the clustered states.
   %BJC2: I'm unsure of the precise reasoning here.
  We could not simply take the mean of the clustered $A$ and $\mathbf{e}_t$ 
  parameters because it is not clear that this is a linear 
  operation for this problem. To circumvent this problem, we modified the ($r$,
  $z$) solute trajectories so that they had a mean of zero, leaving only the 
  fluctuations. We did this by subtracting the mean from each same-state segment of the 
  unclustered trajectory. We used the IHMM algorithm on this modified trajectory
  to infer the clustered state parameters by fixing the clustered state sequence.

@@

  Finally, we generated stochastic trajectory realizations using the clustered
  parameter sets. We drew state sequences with transition probabilities given by
  $T$. While in a given state, we simulated motion according to the VAR(1)
  parameterization of that state. After each state transition, we reset the 
  unconditional mean of each state based on the particle's position immediately
  before the state transition occurred.

@@

@@

  \subsection{Reproducing MD Trajectories and MSDs with the IHMM}

>>>>

\documentclass[journal=jpcbfk,manuscript=article]{achemso}

\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{subcaption}
\usepackage{amsmath} % or simply amstext
\usepackage{amssymb}
\usepackage{siunitx}
\usepackage{booktabs}
\usepackage[export]{adjustbox}
\newcommand{\angstrom}{\textup{\AA}}
\usepackage{cleveref}
\usepackage{booktabs}
\usepackage{gensymb}
\usepackage{float}
\usepackage{xr}
%\DeclareUnicodeCharacter{03C9}{$\omega$}
\externaldocument[S-]{Supporting_Information}

\title{Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series 
       of Solute Trajectories in Nanostructured Membranes}
\author{Benjamin J. Coscia}
\affiliation{Department of Chemical and Biological Engineering, University of Colorado Boulder, Boulder, CO 80309, USA}
\author{Christopher P. Calderon}
\affiliation{Department of Chemical and Biological Engineering, University of Colorado Boulder, Boulder, CO 80309, USA}
\author{Michael R. Shirts}
\email{michael.shirts@colorado.edu}
\affiliation{Department of Chemical and Biological Engineering, University of Colorado Boulder, Boulder, CO 80309, USA}

@@

  \begin{abstract}

  Appropriate time series modeling of complex diffusion on the microsecond time scale
  can provide a path towards inferring transport mechanisms and predicting bulk properties
  characteristic of much longer time scales. In this work we apply the infinite hidden 
  Markov model (IHMM) to solute center-of-mass trajectories generated from long 
  molecular dynamics simulations in a cross-linked inverted hexagonal phase lyotropic 
  liquid crystal (LLC) membrane in order to automatically detect and parameterize an unknown 
  number of distinct latent vector autoregressive dynamical modes.

  We group similar solute behavior into clusters, analyze the dynamics characteristic
  to each cluster and relate them to physical interactions between solutes and the 
  membrane. This enhancement to our understanding of the diverse range of influential 
  solute-membrane interactions allows us hypothesize design changes to LLC monomers 
  aimed towards controlling the rates of solute passage, thus improving the selective 
  performance of LLC membranes.

  Along with parameters of individual states, the IHMM simultaneously infers a transition
  matrix which allows us to stochastically propagate solute behavior from all of the 
  independent trajectories onto much longer time scales while still preserving the 
  qualitative behavior characteristic of the MD trajectories. This affords a direct 
  connection to important macroscopic observables used to characterize performance like
  solute flux and selectivity.

  Overall, this work provides an effective way to simultaneously identify transport 
  mechanisms and project complex diffusive behavior on long time scales.

  \end{abstract}

@@

  % BJC: Not sure if this should be the first paragraph, or if I could just
  % delete this paragraph and start with the next one. 
  There is a need for highly selective membranes in order to perform efficient 
  separations of the components of complex aqueous streams.
  \begin{itemize}
    \item Desalination and boric acid removal from seawater.
    \item Organic micropollutants
    \item Hydraulic fracturing flowback water
    \item Breathable chemical warfare agent barriers
    \item High purity chemicals
    \item While many researchers focus on membrane permeability, we may be 
    able to reduce costs of commercial nanofiltration and reverse osmosis with
    higher selectivity.~\cite{werber_materials_2016}
  \end{itemize}

  The ability to efficiently design novel membrane materials for selective separations would be 
  greatly enhanced by easily interpretable methods for connecting microscopic fluctuations
  with macroscopic observables.
  \begin{itemize}
  	\item If one could apply molecular dynamics (MD) simulations to explain important
  	experimental metrics such as solute flux and selectivity in terms of detailed 
  	chemically dependent solute motion, then we can more easily bridge the disconnect
  	between theory and application.
  	%BJC: this could potentially be moved further down
  	\item There are two main challenges associated with this goal.
  	%since complex and
  	%diverse diffusive motion on time scales accessible to molecular simulations prevents
  	%us from easily relating chemically dependent solute motion to experimental time scales.
  	\item First, we need a simple way to identify and parameterize the fluctuations
  	characteristic to dominant solute-membrane interactions.
  	\item Second, we need a way to propagate this behavior far beyond the time scales
  	afforded by MD.
  \end{itemize}

  In this work, we are particularly interested in the design of lyotropic liquid crystal (LLC)
  monomers, a class of amphiphilic molecules whose ordered self-assembled phases can be 
  cross-linked into mechanically strong membranes capable of highly selective separations.
  \begin{itemize}
  	\item Inverted hexagonal, H\textsubscript{II}, phase LLC membranes are characterized
  	by hexagonally packed, uniform-sized and straight pores, an ideal geometry for high 
  	throughput transport.
	\item Their pores are lined with the LLC monomer functional groups which can
	potentially be designed to interact with solutes in a chemically-specific manner.
	\item It would be extremely useful to experimentalists if we could connect LLC monomer
	design with measurable quantities.
  \end{itemize}

  We have already progressed our understanding of LLC monomer design in our previous work
  by building stochastic time series models based on the chemical intuition gained
  from qualitative MD studies of solute transport in an H\textsubscript{II} phase LLC
  membrane.~\cite{coscia_chemically_2019,coscia_capturing_2020}
  \begin{itemize}
    \item We identified three classes of trapping behavior which result in short time scale
    subdiffusion.
    \item In our first approach to stochastically model this behavior, we treated the
    system's dynamics as a sequence of anti-correlated hops between power law distributed
    periods of entrapment, a framework called subordinated fractional Brownian or L\'evy
    motion.
    \item In a second approach, we treated solute motion as a Markov state model with 
    state-dependent dynamics, where we parameterized the state transition probabilities 
    between each of eight discrete states, defined by the observed trapping mechanism, 
    as well as the solute dynamics within each of these states. 
    \item We showed how one could use realizations of any stochastic model like these in
    order to predict macroscopic flux and selectivity. 
  \end{itemize}

  Although powerful and insightful, our previous work required considerable human effort
  in order to analyze the MD trajectories and to hypothesize stochastic models which
  matched the behavior of solutes.
  \begin{itemize}
    \item It would be significantly more useful and broadly applicable if we could 
    design an approach which automatically distinguishes and parameterizes varied solute 
    dynamics and leaves a facile way to generate an ensemble of stochastic trajectory
    realizations that can be used to predict macroscopic transport properties.
  \end{itemize}

  In this work we attempt to minimize human effort by applying the infinite hidden 
  Markov Model (IHMM)~\cite{beal_infinite_2002,calderon_inferring_2015}, a nonparameteric 
  Bayesian classification technique, in order to automatically detect and infer the 
  parameters of an unknown number of latent vector autoregressive (VAR) modes present 
  in solute center-of-mass time series trajectories. 
  \begin{itemize}
    \item The IHMM has not been used much for MD trajectory analysis.
    \item It's been used extensively for speaker diarization and has found some application
    in single particle trajectory analysis.
  \end{itemize}

  We aim to address two primary scientific 
  questions with this work. 
  %BJC: this may be redundant from the second paragraph. Maybe better suited here?
  First, we want
  to know if the IHMM algorithm can help researchers efficiently uncover underlying
  transport mechanisms which give rise to different solute dynamical behavior.
  Second, we want to evaluate the ability of our model to reproduce solute trajectories
  in a stochastic manner so that we can use it to predict macroscopic transport 
  properties which would otherwise require much longer MD simulations.

@@

  Finally, we use the VAR parameters and state transition probability matrix in
  order to generate stochastic trajectory realizations with behavior similar to that
  exhibited by MD. 
  \begin{itemize}  
    \item One can project these realizations onto much longer timescales with 
    computational ease and use them to predict macroscopic properties.
  \end{itemize}

%  Because independent solute trajectories generated by MD simulations share 
%  similar state dynamics, we can use the IHMM to generate stochastic trajectory 
%  realizations which sample from all behavior exhibited by MD. We verify that our
%  stochastic realizations reproduce the expected hopping and trapping behavior 
%  exhibited by MD. We quantitatively validate these trajectories by comparing 
%  their mean squared displacement (MSD) to those measured by MD.

%
%%  \noindent Amphiphilic molecules are capable of self-assembling into ordered nanostructures.
%%  \begin{itemize}
%%    \item 
%%  \end{itemize}
%
%  There are limits to what we can learn from experiment about LLC membrane design.
%  \begin{itemize}
%    \item Experimental observables like permeability and selectivity allow us to speculate
%    about the molecular origins of separation processes.
%    %MRS2: not quite sure what you mean below?
%    \item This drives an empirical design approach which can potentially neglect key
%    interactions which influence selectivity.
%    \item LLC membranes have been shown to exhibit selectivities which cannot be fully 
%    explained by relatively simple macroscopic models. 
%  \end{itemize}
%  
%  Molecular Dynamics (MD) simulations can give us mechanistic insights with atomistic 
%  resolution so that we can intelligently design new membranes for solute-specific 
%  separations.
%  \begin{itemize}
%    \item In our previous work, we built a detailed atomistic model which we used
%    to understand the nanoscopic structure of an LLC Membrane.~\cite{coscia_understanding_2019}
%    \item We also used the model in order to gain a qualitative understanding of 
%    trapping mechanisms which lead to subdiffusive transport behavior.~\cite{coscia_chemically_2019}
%  \end{itemize}
%
%  Unfortunately, the timescales that we can simulate with MD are insufficient to be
%  able to make well-converged predictions of macroscopic transport properties 
%  traditionally used to characterize membranes in the lab.
%  \begin{itemize}
%    %MRS2: would be good to get this next point into the first line of a paragraph, since it's quite important. 
%    \item However, if we use descriptive stochastic models that can capture solute
%    dynamics, then we could project long timescale behavior in addition to gaining
%    a deeper understanding of solute behavior on short timescales. 
%  \end{itemize}

@@

  Using the IHMM framework, we estimated the most likely sequence of hidden states in
  each solute center-of-mass trajectory while simultaneously inferring VAR(1)
  parameters for each state and the overall state transition probability matrix, $T$.
  We created a python implementation of this process which we heavily adapted from
  %BJC: fox_bayesian is citation Emily Fox asked me to use.
  the MATLAB code of Fox et al.~\cite{fox_bayesian_2010} Parameter estimation is iterative. 
  Therefore, we looked for convergence of each entry in the $\mathbf{c}$ vector as well
  as the $A$ and $\Sigma$ matrices of Equation~\ref{eqn:var}. We detected equilibration 
  of the parameters using the module \texttt{pymbar.timeseries.detect\_equilibration}~\cite{chodera_simple_2016} on
  the time series of parameter estimates. Plots illustrating parameter convergence are 
  available in Section~\ref{S-section:convergence} of the Supporting Information. We 
  refer the interested reader to much more extensive descriptions of the inference and 
  sampling procedures used to estimate the VAR(1) parameters and the state sequence. ~\cite{beal_infinite_2002,teh_hierarchical_2006,van_gael_beam_2008,fox_nonparametric_2009,fox_bayesian_2010}

  %BJC: This paragraph could probably be cut down quite a bit. Leave justification for SI
  We carefully applied the IHMM algorithm in a way which takes advantage of the
  system's cylindrical symmetry. While there
  %There 
  are a number of viable ways in which one could choose to analyze these time series,
  %BJC: cutting down on verbosity
  % We could use all 3 dimensions straight
  %from the output trajectory. We could use just the $z$ dimensional coordinate 
  %since we are primarily interested in through-plane membrane transport. Ultimately, 
  we chose to use a multi-step procedure that we believe adequately distinguishes 
  every type of distinct dynamical behavior exhibited by the solutes. The final
  parameterization of the model is in cylindrical coordinates since each pore is a
  cylinder where we expect solutes to exhibit radially symmetric dynamics. We 
  present a step-by-step demonstration of the procedure that follows in 
  Section~\ref{S-section:ihmm_procedure} of the Supporting Information.

@@

  We applied the IHMM to each of the 24 solute trajectories independently.
  Although the IHMM is capable of identifying an infinite number of states, 
  a Dirichlet process tends to exhibit a ``rich get richer" effect, favoring
  a fewer number of states.~\cite{dreyer_discovering_2011} By applying the algorithm to each trajectory 
  independently, we reduce the possibility of lumping together multiple 
  similar states which we would prefer to stay separated before clustering.
  Note that the states identified by the IHMM are heavily influenced by 
  the Gaussian prior placed on $\mathbf{c}$ in Equation~\ref{eqn:var}. 
  In Section~\ref{S-section:prior_guesses} of the Supporting Information we 
  outline our method for choosing reliable prior parameters. 
%  The entries of $A$ and $\Sigma$ do 
%  not vary over a wide range, so the final parameters were relatively insensitive
%  to their priors.

  We ran 2000 iterations of the IHMM procedure in order to arrive at converged 
  state sequences. We evaluated the convergence of the state sequence based on
  the convergence of the VAR parameters. The parameters of each state
  generally converge within 500--1000 iterations. However, the boundaries of 
  each state segment fluctuate which can lead to high variance in the $A$
  and $\Sigma$ parameters in segments with a relatively low number of emissions.
  In these cases, the mean in each dimension tends to be a more reliable indicator of
  convergence (see Figure~\ref{S-fig:convergence3d} of the Supporting Information).

@@

  We defined the finalized parameters of each state as the average of the parameters 
  from each iteration recorded after the equilibration time point. We found the 
  equilibration time point of each component of the $\mathbf{c}$ vector as well as 
  $A$ and $\Sigma$ matrices, and then used the longest equilibration time of all 
  dimensions as the equilibration time point.

  %BJC: I'll need to expand on this a bit to make it clear. Talk about the bootstrapping etc.
  Finally, we generated stochastic trajectory realizations using the finalized
  parameter sets. We drew state sequences with transition probabilities given by
  $T$. While in a given state, we simulated motion according to the VAR(1)
  parameterization of that state. After each state transition, we reset the 
  unconditional mean of each state based on the particle's position immediately
  before the state transition occurred.

  \subsection{Parameter Clustering}\label{method:clustering}

@@

  We reduced the parameter space used agglomerative clustering, a hierarchical
  clustering approach which uses a linkage criteria in order to successively merge
  similar clusters until a desired intracluster distance threshold or number of
  clusters is reached.~\cite{pedregosa_scikit-learn_2011} We used the Ward linkage 
  criteria, which works to minimize the sum of the squared differences within all
  clusters.~\cite{ward_hierarchical_1963} We elected to choose the number of clusters
  rather than the distance threshold. For our data, non-parametric methods such as 
  Bayesian Gaussian mixture models~\cite{pedregosa_scikit-learn_2011,gelman_bayesian_2013}
  tend to delocalize the clusters in parameter space (see Section~\ref{S-section:agglomerative}
  of the Supporting Information).

  We clustered based on the eigenvalues of $A$ and $\Sigma$, the radial means, $\mu_r$, 
  and the self-transition probabilities, $T_{ii}$, of each state for a total of 6 clustering
  dimensions. The means in $r$ are likely meaningful since the membrane is inhomogeneous, 
  transitioning from hydrophilic pores to hydrophobic tails. To cluster on $T_{ii}$, we first
  cast it in terms of the expected dwell times. The expected value of the number of sequential
  self-transitions is simply $\frac{1}{1 - T_{ii}}$. This relationship correctly implies that 
  dwell times approach infinity as $T_{ii}$ approaches 1. There we found the most success with
  agglomerative clustering by linearizing this relationship and clustering on $-log(1 - T_{ii})$
  (see Section\ref{S-section:selfT_cluster} of the Supporting Information for a graphical
  illustration).

%  One can choose alternative clustering
%  features, such as the eigenvalues of the $A$ and $\Sigma$ matrices, but we chose their 
%  diagonals because they can be easily translated to solute behavior. The diagonals of $A$
%  describe the correlation between sequential hops while the diagonals of $\Sigma$ describe
%  the size of the hops in each dimension. We assume that cross correlations in the parameters
%  are negligible for clustering.
%   
%  We also chose to cluster the diagonals of each matrix and the radial means
%  independently before combining the clusters to assign final state labels. If we divide
%  the total states into $m$ clusters based on the diagonal entries of $A$, $n$ more clusters 
%  based on the diagonal entries of $\Sigma$, and $k$ clusters based on $\mu_r$, then there
%  are a total of $mnk$ possible cluster combinations. This allows slightly more flexibility
%  in the total number of states compared to clustering on all five features at once, since 
%  there can be up to, but not necessarily exactly, $mnk$ states.

  %BJC: to be reworked after I have a working procedure.
  Choosing the number of clusters can be somewhat subjective so
  we attempted to add some structure to the selection process by following
  a set of qualitative and quantitative guidelines. We used the silhouette 
  test in order to score the quality of clustering as a function of the number 
  of clusters chosen.~\cite{kaufman_finding_2009} For our data, the silhouette test generally favors the 
  lowest number of clusters possible (see Section~\ref{S-section:nclusters} of
  the Supporting Information). However, choosing too few clusters tends to 
  not distinguish between visually obvious differences in 
  dynamic behavior. This results in finalized parameter sets that are averages of
  distinct behavior which presents further problems with the predictive modeling
  that we discuss later on. We aimed to maintain the highest silhouette score,
  and thus lowest number of clusters, possible while verifying that visually 
  distinct states stayed separated. Based on these guidelines, we decided to 
  group the $A$ and $\Sigma$ parameters into five clusters each and the radial
  means into three clusters. Of the 75 distinct states allowed by this formulation,
  the solutes in this study showed behavior from 32--36 distinct clusters. While
  this may seem like a high number of states, many are sampled infrequently.

  We remapped the state sequence based on the cluster assignments and parameterized
  the clusters using the inference component of the IHMM algorithm. 
%  generated a
%  state transition probability matrix, $T$. The IHMM algorithm also produces an estimate
%  of $T$, but since we fixed the state sequence, we decided to explicitly calculate $T$ 
%  by counting the number of transitions between states compiled across all trajectories.
%  We used the parameter inference component of the IHMM algorithm in order to 
%  infer $A$ and $\Sigma$ of the clustered states.
%  We could not simply take the mean of the clustered $A$ and $\mathbf{e}_t$ 
%  parameters because it is not clear that this is a linear 
%  operation for this problem. To circumvent this problem, 
  First, we modified the ($r$, $z$) solute trajectories so that they had a mean of zero,
  leaving only the fluctuations. We did this by subtracting the mean from each 
  same-state segment of the unclustered trajectory. We used the IHMM algorithm on this
  modified trajectory to infer the clustered state parameters by fixing the clustered 
  state sequence.

@@

  %BJC:This next paragraph can probably be moved
%  Finally, we generated stochastic trajectory realizations using the clustered
%  parameter sets. We drew state sequences with transition probabilities given by
%  $T$. While in a given state, we simulated motion according to the VAR(1)
%  parameterization of that state. After each state transition, we reset the 
%  unconditional mean of each state based on the particle's position immediately
%  before the state transition occurred.

@@

  % BJC: It probably makes sense to put the trajectory generation and MSD predictions
  % first since it will be part of how we justify our clustering choices.

@@

%BJC: will move this earlier
  \subsection{Reproducing MD Trajectories and MSDs with the IHMM}
